---
title: "Datenmengen"
author: "Daniel Zimmermann / David Lilek"
---

# IMPORTANT INFO
The data simulation can't be rendered to html because the ggforest method has to edited manually.

# Data Simulation

```{r import libraries}
library(tidyverse)
library(fitdistrplus)
library(ggplot2)
library(survival)
library(survminer)
```

```{r setup}
#set.seed

set.seed(653)

# Initialize dataframes
# alpha

alpha.hosp <- data.frame(variant=rep("alpha", 764),
                         hosp=TRUE,
                         hosp_er=TRUE)

alpha.er <- data.frame(variant=rep("alpha", 1448-764),
                       hosp=FALSE,
                       hosp_er=TRUE)

alpha.rest <- data.frame(variant=rep("alpha", 34656-1448),
                         hosp=FALSE,
                         hosp_er=FALSE)


# delta

delta.hosp <- data.frame(variant=rep("delta", 196),
                         hosp=TRUE,
                         hosp_er=TRUE)

delta.er <- data.frame(variant=rep("delta", 498-196),
                       hosp=FALSE,
                       hosp_er=TRUE)

delta.rest <- data.frame(variant=rep("delta", 8682-498),
                         hosp=FALSE,
                         hosp_er=FALSE)
```

To take the expected difference in age between hospitalized and non-hospitalized patients into account, we modeled the age as a beta distribution.
The beta distribution is defined between 0 and 1, so when we scale it up by 100, we get a good range for peoples age.
Then, we can adjust the shape parameter for different groups of patients to get different age distributions.

```{r age}
# alpha

age.groups <- data.frame(left=c(rep(0, 2671), # From the paper
                                rep(10, 7373), 
                                rep(20, 6183), 
                                rep(30, 7364), 
                                rep(40, 5588), 
                                rep(50, 3196), 
                                rep(60, 1375), 
                                rep(70, 495), 
                                rep(80, 411)),
                         right=c(rep(10, 2671), 
                                 rep(20, 7373), 
                                 rep(30, 6183), 
                                 rep(40, 7364), 
                                 rep(50, 5588), 
                                 rep(60, 3196), 
                                 rep(70, 1375), 
                                 rep(80, 495), 
                                 rep(NA, 411)))

dist.age <- fitdistcens(age.groups/100, "beta") # Model age distribution
summary(dist.age)
a <- dist.age$estimate["shape1"]
b <- dist.age$estimate["shape2"]

curve(dbeta(x, a, b), from = 0, to = 1)
curve(dbeta(x, a, b*0.5), from = 0, to = 1) # For hospitalized
curve(dbeta(x, a, b*1.2), from = 0, to = 1) # For non-hosp.

alpha.hosp$age <- round(rbeta(nrow(alpha.hosp), a, b*0.5)*100)
alpha.er$age <- round(rbeta(nrow(alpha.er), a, b*0.5)*100)
alpha.rest$age <- round(rbeta(nrow(alpha.rest), a, b*1.2)*100)

# delta

age.groups <- data.frame(left=c(rep(0, 893), 
                                rep(10, 2089), 
                                rep(20, 1453), 
                                rep(30, 1793), 
                                rep(40, 1297), 
                                rep(50, 720), 
                                rep(60, 306), 
                                rep(70, 89), 
                                rep(80, 42)),
                         right=c(rep(10, 893), 
                                 rep(20, 2089), 
                                 rep(30, 1453), 
                                 rep(40, 1793), 
                                 rep(50, 1297), 
                                 rep(60, 720), 
                                 rep(70, 306), 
                                 rep(80, 89), 
                                 rep(NA, 42)))


dist <- fitdistcens(age.groups/100, "beta")
summary(dist)
a <- dist$estimate["shape1"]
b <- dist$estimate["shape2"]

curve(dbeta(x, a, b), from = 0, to = 1)
curve(dbeta(x, a, b*0.5), from = 0, to = 1)
curve(dbeta(x, a, b*1.2), from = 0, to = 1)

delta.hosp$age <- round(rbeta(nrow(delta.hosp), a, b*0.5)*100)
delta.er$age <- round(rbeta(nrow(delta.er), a, b*0.5)*100)
delta.rest$age <- round(rbeta(nrow(delta.rest), a, b*1.2)*100)
```

In the paper, the calendar week has a strong influence, as later hospitalizations might not be registered in the database yet. 
We modeled the cases per week as a geometric distribution (discrete version of exponential)
For alpha, which goes down in later weeks, this can be done directly after shifting the weeks so that the first week of the study (KW13) is defined as 0
For delta, which increases over time, we inverted the time scale to fit the geometric distribution, then changed it back to normal to get the simulated data.

```{r week}
# alpha

alpha.weeks <- c(rep(13, 7593),
                 rep(14, 5568),
                 rep(15, 4673),
                 rep(16, 4431),
                 rep(17, 4058),
                 rep(18, 3608),
                 rep(19, 2608),
                 rep(20, 2177))

dist.week <- fitdist(alpha.weeks-13, "geom") # Model cases per week as geometric distribution
summary(dist.week)
a <- dist.week$estimate["prob"]

plot(c(0:7), dgeom(c(0:7), a))
plot(c(0:7), dgeom(c(0:7), a*1.2)) # Hospitalized
plot(c(0:7), dgeom(c(0:7), a*0.95)) # Non-hospitalized

rgeomt <- function(n, range, prob) { # Truncate distribution to 7 weeks (KW 13 - 20)
  F.a <- pgeom(min(range), prob=prob)
  F.b <- pgeom(max(range), prob=prob)
  
  u <- runif(n, min=F.a, max=F.b)
  
  qgeom(u, prob=prob)
}

alpha.hosp$week <- round(rgeomt(nrow(alpha.hosp), c(0,7), a*1.25)+12)

alpha.er$week <- round(rgeomt(nrow(alpha.er), c(0,7), a*1.05)+12)

alpha.rest$week <- round(rgeomt(nrow(alpha.rest), c(0,7), a*0.95)+12)

# Same for delta, but reversed

delta.weeks <- c(rep(13, 13),
                 rep(14, 67),
                 rep(15, 133),
                 rep(16, 343),
                 rep(17, 632),
                 rep(18, 1377),
                 rep(19, 2144),
                 rep(20, 3973))


dist.week <- fitdist(-delta.weeks+20, "geom")
summary(dist.week)
a <- dist.week$estimate["prob"]

plot(c(7:0), dgeom(c(0:7), a))
plot(c(7:0), dgeom(c(0:7), a*0.8))
plot(c(7:0), dgeom(c(0:7), a*1.05))


delta.hosp$week <- round(-(rgeomt(nrow(delta.hosp), c(0,7), a*0.80)-20))

delta.er$week <- round(-(rgeomt(nrow(delta.er), c(0,7), a*0.95)-20))

delta.rest$week <- round(-(rgeomt(nrow(delta.rest), c(0,7), a*1.05)-20))
```

The number of hospitalized and non-hospitalized patients by vaccination status is given in the paper, so no modeling is needed.

```{r vaccination}
# Numbers directly from paper, no modeling needed
alpha.hosp$vacc <- sample(c(TRUE, FALSE), 
                          nrow(alpha.hosp), 
                          replace=T, 
                          prob=c(228, 536))

alpha.er$vacc <- sample(c(TRUE, FALSE), 
                        nrow(alpha.er), 
                        replace=T, 
                        prob=c(353-228, 1095-536))

alpha.rest$vacc <- sample(c(TRUE, FALSE), 
                          nrow(alpha.rest), 
                          replace=T, 
                          prob=c(6627-353, 28029-1095))


delta.hosp$vacc <- sample(c(TRUE, FALSE), 
                          nrow(delta.hosp), 
                          replace=T, 
                          prob=c(47, 149))

delta.er$vacc <- sample(c(TRUE, FALSE), 
                        nrow(delta.er), 
                        replace=T, 
                        prob=c(129-47, 369-149))

delta.rest$vacc <- sample(c(TRUE, FALSE), 
                          nrow(delta.rest), 
                          replace=T, 
                          prob=c(2001-129, 6681-369))
```

To model the time of hospitalization, we used the Weibull distribution, which is often used for survival rates. 
We assumed an average time from diagnosis to hospitalization of 5-6 days and a right-leaning distribution.
See [article](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7589278/)

```{r combine data}
alpha <- rbind(alpha.hosp, alpha.er, alpha.rest)

# Weibull distribution for modeling survival rates
# On avg. 6 days to hospitalization
# Max. 14 days
# Non-hospitalized set to 14

rweibullt <- function(n, range, shape, scale=1) { # Truncate distribution to 7 weeks (KW 13 - 20)
  F.a <- pweibull(min(range), shape=shape, scale=scale)
  F.b <- pweibull(max(range), shape=shape, scale=scale)
  
  u <- runif(n, min=F.a, max=F.b)
  
  qweibull(u, shape=shape, scale=scale)
}

hosp_er_time <- c(round(rweibullt(1448, c(0,14), 2, 5)), rep(14, 33208)) 

alpha$time <- hosp_er_time


delta <- rbind(delta.hosp, delta.er, delta.rest)

hosp_er_time <- c(round(rweibullt(498, c(0,14), 2, 5)), rep(14, 8682-498))

delta$time <- hosp_er_time


data <- rbind(alpha, delta)
data$variant = as.factor(data$variant)

data <- data[sample(nrow(data)),]

data$age_group <- cut(data$age, 
                      breaks=c(0, 10, 20, 30, 40, 50, 60, 70, 80, 100), 
                      labels=c("0-10", "10-19", "20-29", "30-39", "40-49", 
                               "50-59", "60-69", "70-79", ">80"),
                      include.lowest = TRUE, ordered_result = TRUE, right = FALSE)

str(data)
```


# Visualization 

```{r}

#vaccination status
p <- ggplot(data=data, aes(fill=vacc, x=variant)) +
  geom_bar(position="dodge") +
  ggtitle("Vaccination status at date of specimen")
p
#time of hospitalisation
p <- ggplot(data=data[which(data$hosp_er),], aes(x=time)) +
    geom_bar(position="dodge")
p
# age
p<-ggplot(data, aes(x=age)) +
  geom_density() +
  geom_density(aes(col=hosp))
  theme(legend.position="right") +
  ggtitle("Age Distribution")
p

# calendar week
p <- ggplot(data=data, aes(fill=variant, x=week)) +
  geom_bar() +
  ggtitle("Calendar Week")
p

```


# Cox Regression

* [CoxRegression](http://www.sthda.com/english/wiki/cox-proportional-hazards-model)
* [CoxRegression](https://sphweb.bumc.bu.edu/otlt/mph-modules/bs/bs704_survival/BS704_Survival6.html)
* [CoxModelAssumptions](http://www.sthda.com/english/wiki/cox-model-assumptions)
* [Creating Survival Plots - Cheat Sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/survminer.pdf)
* [CoxRegression nice plotting example](https://rviews.rstudio.com/2017/09/25/survival-analysis-with-r/)

```{r include=FALSE}
# ggforest doesnt work with stratified models
# To fix it, copy the code in the following if statement, then run th cell. 
# Substitute line 10-25 (allTerms <- ...) with the copied code, then save

if (FALSE) {
  ### Copy from here
  allTerms <- lapply(seq_along(terms), function(i) {
    var <- names(terms)[i]
    if(var %in% colnames(data)) {
      if (terms[i] %in% c("factor", "character")) {
        adf <- as.data.frame(table(data[, var]))
        cbind(var = var, adf, pos = 1:nrow(adf))
      }
      else if (terms[i] == "numeric") {
        data.frame(var = var, Var1 = "", Freq = nrow(data), 
                   pos = 1)
      }
      else {
        vars = grep(paste0("^", var, "*."), coef$term, 
                    value = TRUE)
        data.frame(var = vars, Var1 = "", Freq = nrow(data), 
                   pos = seq_along(vars))
      }
    } else {
      message(var, "is not found in data columns, and will be skipped.")
    }
  
  })
  ### To here
}

trace(ggforest, edit=T)

```

## Variant only

For hospitalization we do not see a significant increase in hazard, while for hospitalization + ER we see a moderate hazard ratio (HR) of 1.4.

```{r message=FALSE, warning=FALSE}
#---------------------------------------------------------------
#Cox Regression

res.cox <- coxph(Surv(time,hosp) ~ variant, data = data)
summary(res.cox)
ggforest(res.cox, data = data)

res.cox2 <- coxph(Surv(time,hosp_er) ~ variant, data = data)
summary(res.cox2)
ggforest(res.cox2, data = data)

```

## Variant + Calendar week

Here, we stratified the model by calendar week. This leads to an increased HR. The likely reason is that delta is more prevalent in later weeks, in which the data on hospitalization might not be complete yet. As a result, the HR of delta was underestimated before.

```{r message=FALSE, warning=FALSE}
#---------------------------------------------------------------
res.cox3 <- coxph(Surv(time, hosp) ~ variant + strata(week), data = data)
summary(res.cox3)

ggforest(res.cox3, data = data)


res.cox4 <- coxph(Surv(time,hosp_er) ~ variant + strata(week), data = data)
summary(res.cox4)

ggforest(res.cox4, data = data)
```

## + Vaccination & Week

Including these two additional factors in stratification, we only see small changes to the HR.

```{r message=FALSE, warning=FALSE}
#---------------------------------------------------------------
#Cox Regression
res.cox5 <- coxph(Surv(time, hosp) ~ variant + strata(week) + strata(vacc), data=data)
summary(res.cox5)
# visualize hazard ratio
ggforest(res.cox5, data = data)

res.cox6 <- coxph(Surv(time, hosp_er) ~ variant + strata(week) + strata(vacc), data=data)
summary(res.cox6)
# visualize hazard ratio
ggforest(res.cox6, data = data)
```

## + Vaccination & Week & Age group

```{r message=FALSE, warning=FALSE}
#Cox Regression
#own data right skewed
res.cox7 <- coxph(Surv(time, hosp) ~ variant + strata(week) + strata(vacc) + strata(age_group), data=data)
summary(res.cox7)
# visualize hazard ratio
ggforest(res.cox7, data = data)

res.cox8 <- coxph(Surv(time, hosp_er) ~ variant + strata(week) + strata(vacc) + strata(age_group), data=data)
summary(res.cox8)
# visualize hazard ratio
ggforest(res.cox8, data = data)
```

## Testing assumption

Testing the [Assumptions](http://www.sthda.com/english/wiki/cox-model-assumptions)
p-values should be higher than 0.05 (significance level) --> no significant difference from assumptions

```{r}
test.ph <- cox.zph(res.cox7)
test.ph
ggcoxzph(test.ph)

test.ph2 <- cox.zph(res.cox8)
test.ph2
ggcoxzph(test.ph2)
```

